Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-25 paper/main.tex
working through issues discussed on slack + updated changes doc

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. This change in the number of topics overcame an
undesirable behavior in the UMAP embedding procedure~\citep{McInEtal18a},
whereby embedding coordinates for the 15-topic model tended to be ``clumped''
into separated clusters, rather than forming a smooth trajectory through the 2D
space. When we increased the number of topics to 100, the embedding coordinates
in the 2D space formed a smooth trajectory through the space, with
substantially less clumping (Fig.~\ref{fig:knowledge-maps}). We note that we
used these 2D maps solely for visualization; all relevant comparisons, distance
computations, and statistical tests we report above were carried out in the
original 15-dimensional space, using the 15-topic model. Aside from increasing
the number of topics from 15 to 100, all other procedures and model parameters
were carried over from the preceding analyses. As in our other analyses, we
resampled each lecture's topic trajectory to 1~Hz and projected each question
into a shared text embedding space.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). All participants viewed the videos in the
same order (i.e., \textit{Four Fundamental Forces} followed by \textit{Birth of
Stars}). While we are not aware of any specific confounds of viewing order, nor
have are we aware of how or why viewing order might influence our main
findings, we acknowledge that we did not control for potential order effects in
our study.

We then hand-created 39 multiple-choice questions: 15 about the conceptual
content of \textit{Four Fundamental Forces} (i.e., Lecture~1), 15 about the
conceptual content of \textit{Birth of Stars} (i.e., Lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). One of our group's
undergraduate research assistants worked alongside a rotating Masters student
to develop this set of questions (these researchers are acknowledged in our
paper for their contribution, although they did not meet the criteria for
authorship discussed with all team members at the start of the project, as
determined by J.R.M.) The senior author (J.R.M.) tasked the pair of researchers
with coming up with ``15 conceptual questions about each lecture, along with 9
additional questions about general physics knowledge.'' To help broaden the set
of lecture-specific questions, the researchers were further instructed to work
through each lecture in small segments, identify what each segment was
``about'' conceptually, and then write a question about that concept. The
general physics questions were drawn from the researchers' coursework along
with internet searches and brainstorming with the project team and other
members of J.R.M.'s lab. The final set of questions (and response options) was
reviewed and approved by J.R.M. before we collected or analyzed the text or
experimental data.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions, or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supp.
Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection~\citep[UMAP; ][]{McInEtal18a, McInEtal18b} to construct a 2D
projection of the text embedding space. Sampling the original 100-dimensional
space at high resolution to obtain an adequate set of topic vectors spanning
the embedding space would be computationally intractable. However, sampling a
2D grid is trivial.

>>>>

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. We note that we used these 2D maps solely for
visualization; all relevant comparisons, distance computations, and statistical
tests we report above were carried out in the original 15-dimensional space,
using the 15-topic model. Aside from increasing the number of topics from 15 to
100, all other procedures and model parameters were carried over from the
preceding analyses. As in our other analyses, we resampled each lecture's topic
trajectory to 1~Hz and projected each question into a shared text embedding
space.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). All participants viewed the videos in the
same order (i.e., \textit{Four Fundamental Forces} followed by \textit{Birth of
Stars}).

We then hand-created 39 multiple-choice questions: 15 about the conceptual
content of \textit{Four Fundamental Forces} (i.e., Lecture~1), 15 about the
conceptual content of \textit{Birth of Stars} (i.e., Lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). To help broaden the
set of lecture-specific questions, our team worked through each lecture in
small segments to identify what each segment was ``about'' conceptually, and
then write a question about that concept. The general physics questions were
drawn our team's prior coursework and areas of interest, along with internet
searches and brainstorming with the project team and other members of J.R.M.'s
lab. The final set of questions (and response options) was reviewed and
approved by J.R.M. before we collected or analyzed the text or experimental
data.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions, or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supp.
Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection~\citep[UMAP; ][]{McInEtal18a, McInEtal18b} to construct a 2D
projection of the text embedding space. Whereas our main analyses used a
15-topic embedding space, we used a 100-topic embedding space for these
visualizations. This change in the number of topics overcame an undesirable
behavior in the UMAP embedding procedure~\citep{McInEtal18a}, whereby embedding
coordinates for the 15-topic model tended to be ``clumped'' into separated
clusters, rather than forming a smooth trajectory through the 2D space. When we
increased the number of topics to 100, the embedding coordinates in the 2D
space formed a smooth trajectory through the space, with substantially less
clumping (Fig.~\ref{fig:knowledge-maps}). Creating a ``map'' by
sampling this 100-dimensional space at high resolution to obtain an adequate
set of topic vectors spanning the embedding space would be computationally
intractable. However, sampling a 2D grid is trivial.
